Measuring Hash Trustworthiness via Collision Utility Metrics: Logical Cryptanalysis of MD4

### 2017-12-05 dsn-2018/bare_conf.tex
DSN changes

Evaluating flexibility becomes important as it serves as a
contrapositive to trustworthiness.  A highly flexible attack implies a
larger set of families of collision classes, and decreasing the trust
that should be placed in a hash function. To evaluate how much flexibility exists in a collision,
we propose the following metrics for evaluating the utility of a collision
class:
\begin{enumerate}
    \item The number of unique differentials a collision class has.
    \item The number of unit-step neighbors a collision class has.
    \item The maximum count of zeros in a the binary representation of a
        colliding block (and likewise with ones).
    \item Whether there exists a block which collides under multiple initial
        values.
    \item Whether or not zero, one, or both of the blocks in a collision may
        be of ASCII values under any input block difference.
\end{enumerate}
Note that the first three are quantitative measures providing some measure of
flexibility of a collision class, whereas the latter two are merely looking
for a single witness having such a property. Depending on the scenario,
specific properties of a collision may be of more interest than others.

@@

\subsection{Unjustified Performance Observations}

We share the general tips for working with models of MD4 from a SAT solver
performance perspective, without justification for results. Note that these
results may be limited in usefulness and may not be uniformly applicable.
\begin{itemize}
    \item Unless searching for a second preimage or chosen prefix with
        sufficient number of bits ($\geq 128$), we find models perform better
        leaving the initial state unconstrained.
    \item Adding ASCII block constraints significantly and negatively impact
        the performance of finding a collision.
    \item Keeping a small search radius on neighborhoods involving families
        of collision classes is important for good performance. However,
        once sufficient witnesses have been found across rounds, missing
        families within the round can be found by various filling-in
        techniques through applying the $\precsim$ relation.
\end{itemize}

@@

\textbf{TODO -- Introduction?}

@@

Table \ref{table:differentials} states the number of unique differentials,
including the specified one, which produces a differential path conforming to
each of the collision classes. We note in particular that while Y. Sasaki's
attack is generally an improvement upon X. Wang's, Y. Sasaki's attack has fewer
input block differentials and thus less flexibility in this aspect. Further,
M. Schl{\"a}ffer's attack -- also building off the work of X. Wang -- has the
same number of differentials and is thus also more flexible than Y. Sasaki's.

@@

\subsection{Unit-Step Neighborhood} \label{empirical:neighborhood}

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{Figs/graph-neighborhood-crop.pdf}
\caption{}
\label{Fig:Graph-Neighborhood}
\end{center}
\end{figure}

Refer to Table \ref{table:neighborhood} for the sizes of neighborhoods
of existing collisions. In particular, note that both Y. Sasaki's attack and
M. Schl{\"a}ffer's attack -- which both claimed to improve upon X. Wang's
attack--have larger neighborhood sizes, and likewise with P. Kasselman's
attack which improved upon H. Dobbertin's attack. Thus, the newer collisions
occur in more dense subsets of the collision class space.

@@

@@

See table \ref{table:zeroes} for the maximum count of zeroes in a block
for existing collisions. Note that X. Wang's attack is the leader, with 509
maximum zero bits in the collision pair. Further, while Y. Sasaki improved
the bound in other areas, this was not one of them. Of note too,
H. Dobbertin's and P. Kasselman's collisions are essentially identical in
this metric.

@@

We noted in table \ref{table:neighborhood} that the unit-step neighborhood of
the existing attacks were non-empty. By repeatedly expanding neighborhoods of
known collision classes, we can construct paths in the collision space. We
made the following observations:

>>>>

Evaluating flexibility becomes important as it serves as a
contrapositive to trustworthiness.  A highly flexible attack implies a
larger set of families of collision classes, and decreasing the trust
that should be placed in a hash function. To evaluate how much flexibility exists in a collision,
we propose the following metrics for evaluating the utility of a collision
class:
\begin{enumerate}
    \item The number of unique differentials a collision class has.
    \item The number of unit distance neighbors a collision class has.
    \item The maximum count of zeros in a the binary representation of a
        colliding block (and likewise with ones).
    \item Whether there exists a block which collides under multiple initial
        values.
    \item Whether or not zero, one, or both of the blocks in a collision may
        be of ASCII values under any input block difference.
\end{enumerate}
Note that the first three are quantitative measures providing some measure of
flexibility of a collision class, whereas the latter two are merely looking
for a single witness having such a property. Depending on the scenario,
specific properties of a collision may be of more interest than others.

@@

\subsection{Intuition on Performance}

Our empirical results have provided some intuition on the
performability of our presented attacks, and the general problem of
working with models of MD4 from a SAT solver
performance perspective.  While our results are of a heuristic nature,
and may not be uniformly applicable, we believe they have
generalizable value and provide insight within the specific domain of
collision generation.

\begin{itemize}
    \item Unless searching for a second preimage or chosen prefix with
        sufficient number of bits ($\geq 128$), we find models perform better
        leaving the initial state unconstrained.
    \item Adding ASCII block constraints significantly and negatively impact
        the performance of finding a collision.
    \item Keeping a small search radius on neighborhoods involving families
        of collision classes is important for good performance. However,
        once sufficient witnesses have been found across rounds, missing
        families within the round can be found by various filling-in
        techniques through the application of the $\precsim$ relation.
\end{itemize}

@@

In this section we present our empirical results, evaluating the
utility of existing published collisions by exploring their structure
with 3-CNF-SAT, and building representative neighborhoods using our
distance metrics.  We present these results in the form of an
exploration of unique differentials, exploration of unit distance
neighborhoods, maximum count of zeros in the binary representation of
a colliding block, the existance of blocks which collide under
multiple initial values, and ASCII block collisions as discussed
previously in Section \label{Sec:Intuition}.

@@

Table \ref{table:differentials} illustrates the number of unique differentials,
including the input block differential specified by the original authors, which produces a differential path conforming to
each of the collision classes. We note in particular that while Y. Sasaki's
attack is generally an improvement upon X. Wang's, Y. Sasaki's attack has fewer
input block differentials and thus less flexibility in this aspect. Further,
M. Schl{\"a}ffer's attack -- also building off the work of X. Wang -- has the
same number of differentials and is thus also more flexible than Y. Sasaki's.

@@

\subsection{Unit Distance Neighborhood} \label{empirical:neighborhood}

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{Figs/graph-neighborhood-crop.pdf}
\caption{All collisions in the unit distance neighborhoods of
  Dobbertin and Kasselman collision classes
  organized by connectedness, and showing the inter-distance
  paths neighborhoods of unpublished collisions.}
\label{Fig:Graph-Neighborhood}
\end{center}
\end{figure}

Table \ref{table:neighborhood} presents the sizes discovered empirically of neighborhoods
of existing collisions. In particular, note that both Y. Sasaki's attack and
M. Schl{\"a}ffer's attack -- which both claimed to improve upon X. Wang's
attack--have larger neighborhood sizes, and likewise with P. Kasselman's
attack which improved upon H. Dobbertin's attack. Thus, the newer collisions
occur in more dense subsets of the collision class space.

@@

Figure \ref{Fig:Graph-Neighborhood} graphically illustrates the
geometry of the unit distance neighborhoods for H. Dobbertin and
P. Kasselman's attack, illustrating the geometry of inter-distance
paths within the newly discovered collisions.  While all nodes
represent collisions which are unit-distance from the Dobbertin and
Kasselman collisions, inter-distance paths may be as large as two.
Clear structure emerges in this representation of the cliques, and
corresponds directly to the cliques in Figure \ref{Fig:Graph-Clique}.

@@

We present results in Table \ref{table:zeroes} which represent the maximum count of zeroes in a block
for existing collisions. Note that X. Wang's attack is the leader, with 509
maximum zero bits in the collision pair. Further, while Y. Sasaki improved
the bound in other areas, this was not one of them. Of note too,
H. Dobbertin's and P. Kasselman's collisions are essentially identical in
this metric.

@@

We noted in table \ref{table:neighborhood} that the unit distance neighborhood of
the existing attacks were non-empty. By repeatedly expanding neighborhoods of
known collision classes, we can construct paths in the collision space. We
made the following observations:
